Group polarization replications may be marred by high false discovery rates

### 2025-06-19 GroupPolarizationStatmod/paper/ms.tex
finalized table loc for now
init intro.qmd for only intro drafting

If an extremist opinion falls and it is measured with a Likert scale, will it make a noise?

Unchecked social and political polarization undermines cooperation and
political efficacy \autocite{Mason2018UncivilAgreementBook,Klein2020}.
\emph{Group polarization} is the name given the social phenomenon of 
when a group becomes more extreme in mean opinion over
time, assuming the mean opinion is biased to one extreme to begin with.
\mt{now hit them with the real shit to intro the problem, our analysis, and
how this can be used constructively going forward.}

@@

Experimetnal designs in group polarization studies share a certain
experimental boilerplate (Figure 1 --- NEED TO INCLUDE) with the goal of
testing the group polarization hypothesis that a group of like-minded
people will develop more extreme opinions on some salient topic over the
course of discussing some topic (Figure~\ref{fig:ModelSketch}). 
Group polarization experiments are thus
initialized by forming like-minded groups of people and surveying their
initial opinions on some topic. In some cases these initial opinions are
used for assigning participants to like-minded groups, but other methods
have been used, such as geographic location for producing, on average,
like-minded groups \autocite{Schkade2010}. Participants then discuss the
topic in their groups. After discussion, participant opinions are again
measured. Several studies we analyze here determined group polarization
occurred (i.e., defined a positive detection) only as a non-zero
difference in mean opinion, with no other statistical testing. Some
others performed null-hypothesis \(t\)-tests to test whether there was
enough evidence to reject the null-hypothesis that \emph{simple
consensus} occurredand not group polarization. In our meta-analysis
simulations, we measure simulated group polarization effect size in
terms of Cohen's \(d\), which is an abstraction beyond these two
practices from the group polarization literature that can also be used
to compare model fits between metric models that treat group
polarization measurements as continuous although they are ordinal, and
more appropriate statistical models that account for data ordinality.

>>>>

If an extremist's opinion falls and it is measured with a Likert scale, will it
make a noise? The answer had better be yes if we want to properly measure opinion
change. Accurate measurements are critical
for developing rigorous and reliable methods to promote more centrist,
sustainable, and responsive governments and institutions
\autocite{Mason2018UncivilAgreementBook,Klein2020}. Unfortunately, social and
behavioral scientists have often used statistical methods 
that distort
opinion change when it's measured on an ordinal, Likert-style scale (e.g. -3
indicates ``Strongly disagree'', +3 indicates ``Strongly agree'', and 0 is
neutral)~\autocite{Liddell2018}. 
This includes research that Obama official and law professor Cass Sunstein
marshals to explain why ``people become extremists'' and why ``political and
cultural polarization'' is ``so pervasive in America''\footnote{From the
  Publisher's Summary. Sunstein does not make such general claims in the book,
  but does suggest that the research he reviews collectively provides
  a ``clue'' to explaining how
  ``facism...student radicalism...Islamic terrorism...[t]he Rwandan
genocide...[e]thnic Conflict.}
administration official and law professor Cass Sunstein writing a book called
*researchers construct

\emph{Group polarization} is that essentially
measures how much being in an opinion echo-chamber the name given the social phenomenon of 
when a group becomes more extreme in mean opinion over
time, assuming the mean opinion is biased to one extreme to begin with.
\mt{now hit them with the real shit to intro the problem, our analysis, and
how this can be used constructively going forward.}

@@

The null-consistency procedure functions as a test of model indistinguishability: it demonstrates that a reported group polarization effect could plausibly arise from a stable-latent process combined with changing opinion precision. This logic echoes foundational concerns in causal inference, where two competing models—one causal, one not—can imply the same observed data distribution and are therefore empirically indistinguishable without further assumptions \cite{pearl2000causality,spirtes2000causation}. In such cases, inference from observation to mechanism is invalid. Similarly, in psychometrics, it is well known that different configurations of latent traits and response thresholds can yield identical observed ordinal patterns, making the true structure unidentifiable without additional constraints \cite{borsboom2005measuring}. The present approach does not attempt to estimate the probability of false detection, but rather shows that under ordinal measurement, a null model can reproduce the observed pattern—rendering the result non-diagnostic of a true latent shift. This establishes a stringent standard: an effect must not only be detected, but shown to be incompatible with plausible null-generating mechanisms.

Experimental designs in group polarization studies share a certain
experimental boilerplate (Figure 1 --- NEED TO INCLUDE) with the goal of
testing the group polarization hypothesis that a group of like-minded
people will develop more extreme opinions on some salient topic over the
course of discussing some topic (Figure~\ref{fig:ModelSketch}). 
Group polarization experiments are thus
initialized by forming like-minded groups of people and surveying their
initial opinions on some topic. In some cases these initial opinions are
used for assigning participants to like-minded groups, but other methods
have been used, such as geographic location for producing, on average,
like-minded groups \autocite{Schkade2010}. Participants then discuss the
topic in their groups. After discussion, participant opinions are again
measured. Several studies we analyze here determined group polarization
occurred (i.e., defined a positive detection) only as a non-zero
difference in mean opinion, with no other statistical testing. Some
others performed null-hypothesis \(t\)-tests to test whether there was
enough evidence to reject the null-hypothesis that \emph{simple
consensus} occurredand not group polarization. In our meta-analysis
simulations, we measure simulated group polarization effect size in
terms of Cohen's \(d\), which is an abstraction beyond these two
practices from the group polarization literature that can also be used
to compare model fits between metric models that treat group
polarization measurements as continuous although they are ordinal, and
more appropriate statistical models that account for data ordinality.
